---
title: "Otter Population Analysis Methods and Results"
author: "E Hellwig"
date: "2/1/2018"
output:
  word_document: default
  pdf_document: default
header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage[table]{xcolor}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage[normalem]{ulem}
csl: northeastern-naturalist.csl
bibliography: OtterCitations.bib
---

```{r setup, include=FALSE}
library(ggplot2)
library(rgdal)
library(dismo)
library(randomForest)
library(knitr)
library(kableExtra)
library(rstan)
library(gridExtra)
library(ggmap)
source('../functions.R')
#source('functions.R')
knitr::opts_chunk$set(echo = FALSE)
datapath <- '../../otterData/'


##################
chars <- read.csv(file.path(datapath, 'populationcharacteristics.csv'))
av <- read.csv(file.path(datapath, 'allvars.csv'))

##########For location map
lvlNames <- c('Abbotts Lagoon', 'Alpine Reservoir', 'Bass Lake','Bolinas',
              'Drakes Bay', 'El Estero', 'Giacomini Wetlands','Las Gallinas',
              'Madera Creek', 'Muir Woods','North Tomales Bay','Peters Dam',
              'Rodeo Lagoon','Tennessee Valley')


locs <- readOGR(dsn=file.path(datapath, 'otterSPDF.GeoJSON'),layer="OGRGeoJSON")
names(locs)[c(1,4)] <- c("Latitude", "Region")

bump <- c(-1, 1, 1, 0, 0, 1, 0, 0 ,0, 0, -1, 0, 0, 0) *(-0.007)
tcoords <- coordinates(locs)
tcoords[,1] <- tcoords[,1] + 0.005
tcoords[,2] <- tcoords[,2]-bump

ldf <- locs@data
levels(ldf$Region) <- c('Bay','Coast','Inland','Point Reyes')
levels(ldf$loc) <-lvlNames
ldf$labelX <- tcoords[,1]
ldf$labelY <- tcoords[,2]

ext <- extent(locs)
bbgg <- c(left=xmin(ext), bottom=ymin(ext), right=xmax(ext), top=ymax(ext))
locbasegg <- get_map(location=bbgg, maptype = 'satellite', zoom=10)


#####for big population plot
pp <- read.csv(file.path(datapath, 'vis/popplot.csv')) 
levels(pp$Site) <- lvlNames


b1Q <- read.csv(file.path(datapath, 'beta1quantiles.csv'))


######Locations of otter populations
locs <- readOGR(dsn=file.path(datapath, 'otterSPDF.GeoJSON'),
                layer="OGRGeoJSON")


##full randomforest
fullrf <- readRDS(file.path(datapath, 'fullrandomForest.RDS'))
fullimp <- readRDS(file.path(datapath, 'fullRFvariableimportance.RDS'))



###Moran's I results
mi <- read.csv(file.path(datapath, 'moranI.csv'))
levels(mi$var) <- c('Initial Group Size', "Population Growth Rate", "Decline")
levels(mi$spmethod) <- c('Low', "High")
levels(mi$h1) <- c('Positive','Negative')
####glm analysis
avm <- read.csv(file.path(datapath, 'LatitudeModelResults.csv'))

Amod <- readRDS(file.path(datapath, 'models/AlphaFixedModel.RDS'))
Bmod <- readRDS(file.path(datapath, 'models/BetaFixedModel.RDS'))
DPmod <- readRDS(file.path(datapath, 'models/DeclinePBetaRegModel.RDS'))

Afit <- modfit(Amod)
Bfit <- modfit(Bmod)
DPfit <- modfit(DPmod,  betareg = TRUE)


resid <- read.csv(file.path(datapath, 'Residuals.csv'))
qual <- read.csv(file.path(datapath, 'ParameterEstimates.csv'))


#########################
#color palettes
sitecolors <- c('#4393c3','#252525',"#ca0020")
charcolors <- c('#1b9e77', '#d95f02', '#7570b3')

```


# Methods

The statistical analysis consisted of two parts: a population analysis and a characteristics analysis. All parts of the statistical analyses were conducted in R, and all regression based models were constructed with the 'rstan' package [@stan_art]. These FSS and their regions are shown in Figure 1. The 14 otter Focal Study Sites (FSS) were split between four regions based on expert opinion. Each otter group's home range was classified as primarily estuary, lagoon, reservoir, or stream. FFSs were also assigned the most frequent transportation method used for accessing the site. Potential access methods included car, boat, hiking trail, and no access (Table 1). 


```{r map, fig.cap='Approximate Locations of Focal Study Sites and their corresponding regions.', warning=FALSE}

la <- ggmap(locbasegg, extent='device', legend = "topright") + 
    geom_point(aes(x=Longitude, y=Latitude, color=Region), size=2.5, data=ldf) + 
    geom_text(aes(x=labelX, y=labelY, label=loc), 
              data=ldf, color='white', hjust=0, size=2.5, nudge_x = 0.008) +
    theme(legend.text=element_text(size=8), 
          legend.title = element_text(size=10, face='bold'),
          legend.key.height = unit(5, 'mm'))

la

```



## Population Analysis

The goal of the population analysis was to assess the size and population growth rate of each otter group. We fit a Bayesian mixed effects linear model [@Carpenter2016] to the otter population data times series, using year as the predictor variable and otter group size as the response variable. FSS was included as a varying effect to mitigate the effects of spatial autocorrelation. Next, using the population model, we estimated three population characteristics for each of the sites: initial group size, population growth rate and likelihood that the otter group population is declining. This model structure made it easy to estimate both population growth rate and intial population size for each group. 

We calculated the third model derived characteristic, likelihood of decline, using the posterior distribution of the population growth rate (the coefficient of the predictor variable year). This distribution was generated using Markov Chain Monte Carlo [@mcmc]. The number of samples where the value of the growth rate was less than zero were counted and then divided by the total number of samples. This produced the probability that otter population had a negative growth rate and thus had a declining population.

## Characteristic Analysis

The characteristics analysis focused on investigating correlations between differences in the three site characteristics and other population attributes (Table 1). Due to the large number of potential predictors relative to the number of observations and the strong correlations between attributes, we used a Random Forest analysis to identify the most useful attributes for predicting differences between each of the group characteristic values. Using the 'randomForest' R package [@rf_art], one Random Forest model was fit for each of the three group charateristics using all available attributes as predictors. Then the variable importance measure identified the most important of these variables for predicting differences between otter groups.

```{r char_attribute}

knitr::kable(chars[,-1], format='latex', booktabs=TRUE,
      caption='Otter group variables of interest. ') %>%
    kable_styling(latex_options = 'striped') %>%
    column_spec(3, width="30em")


```


We tested the characteristic data for spatial autocorrelation using Moran's I before fitting regression models using individual population attributes. Moran's I was calculated using the 'spdep' package [@spdep_art]. Both a high and low spatial connectivity structure were used to increase the robustness of the result (Figure 1). We created the high spatial connectivity structure using the three nearest neighbors adjacency scheme and the low connectivity structure using a 10km distance threshold. Next, we fit a fixed effects Bayesian generalized linear models for each of the characteristics using the most important population attribute as a predictor. We used Gaussian likelihoods for both initial population size and population growth. The likelihood of decline necessitated a beta distributed likelihood because it is constrained between zero and one. We then we retested the residuals from those models to check for remaining spatial autocorrelation. 

To assess the quality of the characteristic models, each model was compared to the corresponding intercept only model. We used the root mean squared error (RMSE) as the measure of model quality. RMSEs for all models were calculated using 7-fold crossvalidation.


```{r pop_table, include=TRUE}

regiongrouping <- c(1,5:7,11, 3,4, 10, 13:14, 8:9, 2,12)    
b1Qregroup <- b1Q[regiongrouping, c(4,1,2,3)]

#print(b1Q)

knitr::kable(b1Qregroup, booktabs=TRUE, format='latex', row.names=FALSE,
             col.names=c('Site','Median Growth', '2.5%', '97.5%'),
             caption="Estimates of the population growth parameter, in average otters per year, for each otter group as well as credible intervals for each estimate. Sites are grouped by region.", align='c') %>%
    kable_styling() %>%
    group_rows('Point Reyes', 1, 5)%>%
    group_rows('Coastal', 6, 10)%>%
    group_rows('San Francisco Bay', 11, 12)%>%
    group_rows('Inland', 13, 14)




 
```


# Results

## Population Analysis

Of the 14 otter FSSs, four of them show strong evidence of increasing population sizes: Alpine Reservoir, Giacomini Wetlands, North Tomales Bay and Peters Dam (Figure 1, Table 1). Of the remaining otter sites, three have a greater than 50% likelihood that they are in decline, based on the approximated posterior distribution. For the other seven FSSs, there is not enough data at this time to make any strong conclusions. 


```{r pop_plot, warning=FALSE, fig.cap='Predicted and observed otter population sizes by site. The points indicate observed otter group sizes, while the lines indicate predicted populations.', cache=TRUE}

popp <- ggplot(data=pp) + geom_line(aes(x=Year, y=P_Otters, color=SiteColor),
                                    size=0.7)
popp <- popp + geom_ribbon(aes(x=Year, min=Lower95, max=Upper95), alpha=0.3)
popp <- popp + geom_point(aes(x=Year, y=O_Otters, color=SiteColor), size=1.2)
popp <- popp + facet_wrap(~Site) + theme_bw(10)
popp <- popp + scale_y_continuous(breaks=seq(0, 15, by=5))
popp <- popp + scale_x_continuous(breaks=seq(2013, 2022, by=3))
popp <- popp + scale_color_manual(values=sitecolors, labels=c('Negative', 'Not Significant', 'Positive'), name='Trend Significance')
popp <- popp + coord_cartesian(ylim = c(0, 15)) 
popp <- popp + labs(x='Year', y='Otter Population')


popp

```



## Characteristic Analysis

In the Random Forest analysis, latitude best predicted all three population characteristics (Figure 2). The most of the other predictors that performed nearly as well strongly correlate with latitude in this data set (region, distance to San Francisco, and longitude). Annual visitors was about as predictive as region. Just like region though, it did not add useful information when included in a generalized linear model. Habitat performed notably poorly, along with population density, and primary access method.

Moran's I showed evidence of spatial autocorrelation for both the high and low connectivity spatial structures (Table 3). Initial population size showed strong evidence of spatial autocorrelation using both adjacency schemes. There was a little evidence that likelihood of decline may be spatially autocorrelated as well. The population growth rate did not show evidence of spatial autocorrelation. 

Latitude predicted all three group characteristics better than the equivalent  predictorless model (Table 4). Higher latitudes correlated with higher initial group sizes, larger population growth rates and lower likelihoods of a group being in decline (Figure 3). The was no evidence of spatial autocorrelation in the characteristic regression residuals, indicating that future efforts at modeling these variables should include latitude or a correlated attribute (Table 5). 



```{r random_forest, fig.cap='Random Forest relative predictor importance for initial otter group size, population growth rate, and likelihood of decline.', fig.height=2.5}

vars <- c('Initial Group Size', 'Population Growth Rate',
          'Likelihood of Decline')

rfi <- ggplot(data=fullimp)
rfi <- rfi + geom_point(aes(x=value, y=attribute, color=characteristic))
rfi <- rfi + theme_bw(10) + labs(x="Relative Predictor Importance",
                                 y="Predictor")
rfi <- rfi + scale_color_manual(values=charcolors, 
                                label=vars, name='Characteristics')

rfi
#c('blue2', 'red4', 'salmon4')

#par(mfrow=c(3,1), cex=.7, mar=c(3,2,2,1))
#for (i in 1:length(fullrf)){
 #   varImpPlot(fullrf[[i]], main=vars[i])
#}


```




```{r moranIoriginal, include=TRUE}

#print(mi)

orig_inds <- which(mi$res=='originalData' & mi$h1=='Positive')
#print(orig_inds)

MI_orig <- mi[orig_inds, c('spmethod', 'Moran', 'pval')]
#print(MI_orig)

knitr::kable(MI_orig, format='latex', row.names = FALSE,
      col.names=c('Spatial Connectivity',"Moran's I", 'p-value'), digits=3,
        align='c',
      caption="Values of the Moran's I statistic, a measure of spatial autocorrelation, and their associated pseudo p-values. The low connectivity adjacency scheme used a 10km distance threshold and the high connectivity scheme used the 3 nearest neighbors method.", booktabs=TRUE) %>%
    kable_styling(full_width = FALSE) %>%
    group_rows('Initial Group Size', 1, 2)%>%
    group_rows('Population Growth Rate', 3, 4)%>%
    group_rows('Likelihood of Decline', 5, 6)
    
```



```{r latmodtable}

options(knitr.kable.NA = '')

kable(qual[,-1], format='latex', booktabs=TRUE, digits=2, 
      col.names = c('Coefficient','Estimate','Lower 95%', 
                    'Upper 95%', 'Latitude', 'Null'),
      caption='Coefficient estimates for models using Latitude to predict each of the three group characteristics. Credible intervals for the parameter estimates were computed using the highest posterior density interval (HPDI).') %>%
    kable_styling(latex_options = 'striped') %>%
    collapse_rows(columns=c(1:2), latex_hline = 'none') %>%
    group_rows('Initial Group Size', 1, 2) %>%
    group_rows('Population Growth Rate', 3, 4)%>%
    group_rows('Likelihood of Decline', 5, 6) %>%
    add_header_above(c(" "=2, "HPDI"=2, 'RMSE'=2)) 



```



```{r lat_regs, fig.cap="Relationship between the latitude of the otter group location and all three characteristics: initial group size, population growth rate, and likelihood of decline. The equation in each panel is the linear regression model fitted for each characteristic using a Bayesian framework. The models are visualized by the corresponding colored lines.", include=TRUE, fig.height=7, fig.width=4}

avmA <- avm[avm$response=='alpha', ]
avmB <- avm[avm$response=='beta', ]
avmDP <- avm[avm$response=='declineP', ]

areg <- ggplot() + geom_point(aes(x=Latitude, y=value), 
                              data=avmA[avmA$type=='obs',]) + 
    geom_line(aes(x=Latitude, y=value), color=charcolors[1],
              size=0.9, data=avmA[avmA$type=='fit',]) +
    labs(x='', y='Initial Group Size') + theme_bw() +
    geom_text(aes(x=38, y=3, label=formulastring(Amod)), hjust=0, size=4.5,
              color=charcolors[1])

breg <- ggplot() + geom_point(aes(x=Latitude, y=value), 
                              data=avmB[avmB$type=='obs',]) + 
    geom_line(aes(x=Latitude, y=value),  color=charcolors[2],
              size=0.9, data=avmB[avmB$type=='fit',]) +
    labs(x='', y='Population Growth Rate') + theme_bw() +
    geom_text(aes(x=38, y=-0.25, label=formulastring(Bmod)), hjust=0, size=4.5,
              color=charcolors[2])


dpreg <- ggplot() + geom_point(aes(x=Latitude, y=value), 
                              data=avmDP[avmDP$type=='obs',]) + 
    geom_line(aes(x=Latitude, y=value),  color=charcolors[3],
              size=0.9, data=avmDP[avmDP$type=='fit',]) +
    labs(x='Latitude', y='Likelihood of Decline') + theme_bw() +
    geom_text(aes(x=38, y=0.75, label=formulastring(DPmod, beta=TRUE)),
              hjust=0, size=4.5,  color=charcolors[3], parse=T)

gA <- ggplotGrob(areg)
gB <- ggplotGrob(breg)
gDP <- ggplotGrob(dpreg)

allregs <- arrangeGrob(
  gA, gB, gDP, nrow = 3, ncol=1, widths=0.5, heights = rep(6, 3))

plot(allregs)

```




```{r newresid_moran}


MI_res <- mi[mi$res=='residuals', c('spmethod','h1', 'Moran', 'pval')]

knitr::kable(MI_res, format='latex', row.names = FALSE,
      col.names=c('Spatial Connectivity','Autocorrelation',
                  "Moran's I", 'p-value'), digits=3, align='c',
      caption="Values of the Moran's I statistic for the characteristic regression residuals, and their associated pseudo p-values for both positive and negative autocorrelation. The low connectivity adjacency scheme used a 10km distance threshold and the high connectivity scheme used the 3 nearest neighbors method.", booktabs=TRUE) %>%
    kable_styling(full_width = FALSE) %>%
    group_rows('Initial Group Size', 1, 4)%>%
    group_rows('Population Growth Rate', 5, 8)%>%
    group_rows('Likelihood of Decline', 9, 12)

```

\newpage

# References